---
title: "Absolute quantification with calibration standards"
author: "Matthew J. Smith"
date: "`r Sys.Date()`"
package: quantMSImageR
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Absolute quantification with calibration standards}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE,
                      message = FALSE)

# Gate the evaluated chunks on the presence of the bundled synthetic cal data.
cal_dir  <- system.file("extdata", "cal_example.raw", package = "quantMSImageR")
have_cal <- nzchar(cal_dir) && file.exists(file.path(cal_dir, "cal_MSI.RDS"))
```

# Overview

When a calibration standards acquisition is available, `quantMSImageR` fits a
response-versus-amount curve per lipid and converts tissue-pixel intensities to
`pg/pixel` and `pg/mm2`. The workflow is three steps:

1. `summarise_cal_levels()` -- average each calibration level per standard spot,
2. `create_cal_curve()` -- fit the response-vs-amount curve per lipid,
3. `int2conc()` -- apply the curves to convert tissue-pixel intensities to
   concentration.

This vignette demonstrates the manual API on the small synthetic calibration
dataset bundled with the package, then shows how the same steps are driven from
a study YAML configuration. `run_example(calibrate = TRUE)` runs the whole chain
end-to-end on the bundled data. The core imaging workflow (SNR filtering, ion
images, heatmaps) is covered in the *Introduction to quantMSImageR* vignette.

# The bundled calibration dataset

The synthetic calibration object (generated by `inst/generate_cal_data.R`) is a
`quant_MSImagingExperiment` whose pixels are labelled `Cal`, `Tissue` or `Noise`
in `sample_type`, with a unique `identifier` per calibration spot. Its companion
metadata table maps each `identifier` to a `lipid`, `amount_pg` and `level`.

```{r load, eval = have_cal}
library(quantMSImageR)
library(ggplot2)

cal_dir <- system.file("extdata", "cal_example.raw", package = "quantMSImageR")
cal <- as(readRDS(file.path(cal_dir, "cal_MSI.RDS")),
          "quant_MSImagingExperiment")
cal_metadata <- read.csv(file.path(cal_dir, "calibration_metadata.csv"))

table(pData(cal)$sample_type)
head(cal_metadata)
```

# Fitting the calibration curves

`summarise_cal_levels()` averages the signal at each calibration spot, and
`create_cal_curve()` fits a linear response-vs-amount model per lipid (here
`cal_type = "cal"` for on-slide standards; use `"std_addition"` for on-tissue
standard addition, which subtracts a background level first).

```{r fit, eval = have_cal}
cal <- summarise_cal_levels(cal, cal_metadata, val_slot = "intensity",
                            cal_label = "Cal", id = "identifier")
cal <- create_cal_curve(cal, cal_type = "cal")

# R-squared of each per-lipid curve
cal@calibrationInfo@r2_df
```

The summarised response-per-pixel values and the fitted line for each lipid --
the calibration curves:

```{r curve-plot, eval = have_cal, fig.wide = TRUE}
cal_data <- cal@calibrationInfo@cal_response_data

ggplot(cal_data, aes(pg_perpixel, response_perpixel)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +
  facet_wrap(~ lipid, scales = "free") +
  labs(x = "amount (pg / pixel)", y = "response per pixel",
       title = "Calibration curves")
```

# Converting tissue pixels to concentration

`int2conc()` inverts each fitted curve to turn tissue-pixel intensities into
`pg/pixel`, and uses the acquisition's pixel size to also report `pg/mm2`.

```{r conc, eval = have_cal}
cal <- int2conc(cal, val_slot = "intensity", pixels = "Tissue")
range(spectra(cal, "conc - pg/mm2"), na.rm = TRUE)
```

Each tissue pixel's intensity is mapped onto the calibration curve to give a
concentration -- shown here for the first lipid:

```{r conc-plot, eval = have_cal, fig.small = TRUE}
tissue <- pData(cal)$sample_type == "Tissue"
lipid1 <- fData(cal)$name[1]

df <- data.frame(
  intensity = as.numeric(spectra(cal, "intensity")[1, tissue]),
  pg_pixel  = as.numeric(spectra(cal, "conc - pg/pixel")[1, tissue])
)

ggplot(df, aes(intensity, pg_pixel)) +
  geom_point(alpha = 0.5) +
  labs(x = "intensity", y = "concentration (pg / pixel)",
       title = paste("Intensity to concentration:", lipid1))
```

# Driving calibration from a study YAML

In a full study, calibration is enabled through the `calibration:` block of the
configuration file (`system.file("config_template.yaml", package =
"quantMSImageR")`), positioned just before the `ratios:` block:

```yaml
calibration:
  enabled: true
  cal_acquisition: "cal_run_slide1"     # standards .raw acquisition
  cal_roi_csv:     "cal_ROIs.csv"       # pixel -> sample_type = "Cal" + identifier
  cal_metadata:    "calibration_metadata.csv"
  cal_type:        "std_addition"       # std_addition | cal
  val_slot:        "intensity"
  background_level: "background"
  quantify_pixels: ["tissue_pixels"]
```

With `enabled: true`, `run_study.R` builds the curves from the standards
acquisition and writes a `<study>_calibrated.RDS` alongside the reports:

```{r run-study, eval = FALSE}
CONFIG_FILE <- "path/to/study_config.yaml"
source(system.file("run_study.R", package = "quantMSImageR"))
```

# Session info

```{r session}
sessionInfo()
```
