---
title: "Introduction to quantMSImageR"
author: "Matthew J. Smith"
date: "`r Sys.Date()`"
package: quantMSImageR
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{1. Introduction to quantMSImageR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE,
                      message = FALSE)
```

# Introduction

## The problem

A targeted DESI-MRM imaging run produces one intensity per transition per pixel,
across tens of thousands of pixels and several acquisitions. Turning that into a
biological result runs into three problems at once.

**Most pixels are not tissue.** A section occupies part of the scanned area, and
the surrounding slide still returns signal -- chemical background, carryover and
detector noise. Left in, those pixels drag down every summary statistic and make
images look more uniform than they are. Which pixels are tissue has to be
decided per acquisition, and the decision has to survive being re-applied to a
different MRM panel of the same sample.

**Not every monitored transition yields usable analyte signal.** A transition
can be listed in the method and still carry no usable signal in a given section,
or sit so close to background that it is indistinguishable from it. Reporting
those alongside genuine measurements is how spurious hits reach a figure.

**Response is not amount.** Response depends on ionisation efficiency and local
matrix suppression, so raw responses should not be interpreted as analyte
amount, and their comparability between analytes, sections or instruments should
not be assumed without appropriate normalisation and calibration.

## What this package does

`quantMSImageR` addresses each of those: signal-to-noise (SNR) filtering against
background pixels, tissue/background separation, per-feature ion images and
quantile heatmaps -- and **quantification** against calibration standards for
the third.

## Scope

The package is currently validated for **Waters DESI-MRM** data. Other targeted
imaging data may be analysed after conversion to a compatible
`MSImagingExperiment`, but those workflows have not yet been formally validated.

Within that scope it is designed to handle:

- background-referenced filtering against non-tissue pixels;
- targeted ion-image visualisation and `.txt` export;
- cross-sample summaries such as quantile heatmaps;
- calibration-based amount estimation.

It does **not** do the following, by design -- other tools do them better:

| Not provided | Use instead |
|---|---|
| Untargeted peak detection | a general MSI processing package |
| Feature annotation / molecular identification | dedicated annotation tools |
| Vendor-neutral raw-data conversion | vendor or community converters |
| Histological registration | [TissUUmaps](https://tissuumaps.github.io/); our own helper tools are still in development ([PMID 42438167](https://pubmed.ncbi.nlm.nih.gov/42438167/)) |
| General-purpose image segmentation | image-analysis packages |
| Complex spatial / colocalisation analysis | [corrMSI](https://github.com/MJS-708/corrMSI/tree/dev) |

## The pipeline

Functions are grouped into families, and each section below opens with the
family it belongs to. The whole pipeline is:

```
  .raw acquisition            ion library CSV
         |                          |
         +--------> read_mrm() <----+
                        |  MSImagingExperiment, fData joined to library
                        v
        select_tissue_pixels() --> tissue_pixels.csv  (x, y, tissue/background)
                        |
    +-------------------+-------------------+
    |  ASSEMBLE   setCommonAxis                                  |
    |             align_features -> combine_MSIs / bind_panels   |
    |             trim_MSI                                       |
    +-------------------+-------------------+
                        v   quant_MSImagingExperiment
    +-------------------+-------------------+
    |  FILTER     int2response  (IS-normalise)   |
    |             zero2na, remove_blank_mzs      |
    |             int2snr -> applySNR, back2NA   |
    +-------------------+-------------------+
                        v
         +--------------+--------------+
         v                             v
   VISUALISE                     QUANTIFY  (needs a standards acquisition)
   imageR                        summarise_cal_levels
   quantile_hm                          v
                                 create_cal_curve
                                        v
                                 int2conc -> "pg_pixel", "pg_mm2"
                                        v
                                 plot_cal_coverage   (does calibration cover the data?)

  WORKFLOW:  run_study() (YAML)  .  generate_txt_images()  .  run_example()
```

Quantification is covered in the companion vignette; everything else is below.

Per-feature images can also be exported as plain `.txt` intensity matrices,
which load directly into ImageJ and similar tools -- useful for downstream
multimodal analysis such as co-registration with histology or other imaging
modalities.

Quantification against calibration standards is covered separately in the
*Quantification with calibration standards* vignette
(`vignette("quantification", package = "quantMSImageR")`).

The workflow implemented here is described and applied in
[Smith *et al.*, *Analytical Chemistry* (2024)](https://pubs.acs.org/doi/10.1021/acs.analchem.4c02350).

The package builds on
[Cardinal](https://bioconductor.org/packages/Cardinal/): every acquisition is a
`Cardinal::MSImagingExperiment`, and most `quantMSImageR` functions operate on
the `quant_MSImagingExperiment` subclass, so we coerce first.

# Installation

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("quantMSImageR")
```

# Quick start

The fastest way to see the whole pipeline is `run_example()`, which runs the
imaging workflow on two synthetic groups (**A**, circular tissue, n = 3;
**B**, square tissue, n = 3), renders the HTML report, and also demonstrates
quantification on the bundled calibration standards:

```{r quickstart, eval = FALSE}
library(quantMSImageR)
res <- run_example()          # both groups; imaging report + calibration demo
res$calibrated                # the calibrated object
```

The rest of this vignette reproduces the key steps by hand.

# The data model

Every acquisition is a `Cardinal::MSImagingExperiment`. `quantMSImageR` adds the
`quant_MSImagingExperiment` subclass, which carries two extra slots for
calibration and tissue metadata, so most functions coerce with `as()` first.

The object is **features x pixels**: one row per MRM transition, one column per
pixel. Calculated quantities -- `response`, `snr`, calibrated amount -- are
stored as *additional* spectra layers. Masking functions are the exception:
`applySNR()`, `back2NA()` and `zero2na()` modify the selected `val_slot` in
place, so keep a copy of the object when the unmodified values are needed
afterwards.

| Component | Contains | Accessor |
|---|---|---|
| spectra slots | `intensity` (raw), `response` (IS-normalised), `snr`, `pg_pixel` / `pg_mm2` (calibrated) | `spectra(x, "name")`, `spectraData(x)` |
| feature metadata | `mz`, `name`, transition m/z, analyte type, ion-library columns | `fData(x)` |
| pixel metadata | `x`, `y`, `run`, tissue/background label | `pData(x)` |
| calibration | fitted models, calibration levels, fit diagnostics | `calibrationModels(x)`, `calibrationLevels(x)`, `calibrationDiagnostics(x)` |
| tissue summaries | pixel and ROI matrices | `tissueMatrix(x)`, `tissueData(x)` |

Multiple acquisitions are held in **one object**, distinguished by the `run`
column of `pData()` -- that is what `combine_MSIs()` produces and what every
per-sample summary groups on.

Pixel metadata uses a consistent vocabulary: `run` identifies the acquisition,
`x`/`y` are grid coordinates, and the tissue mask lives in `sample_name` as
`tissue_pixels` / `background_pixels`. (Masks written by earlier versions used
`noise_pixels`; both spellings are still accepted.) The calibration path uses
`sample_type` with `Cal` / `Tissue` / `Background`.

```{r data-model}
library(quantMSImageR)

p   <- system.file("extdata", "example.raw", "section01.RDS",
                   package = "quantMSImageR")
obj <- as(readRDS(p), "quant_MSImagingExperiment")

dim(obj)                        # features x pixels
names(spectraData(obj))         # available layers
head(as.data.frame(fData(obj)), 3)
head(as.data.frame(pData(obj)), 3)
```

# Reading your own data

The bundled example ships as `.RDS` so the vignette is reproducible, but real
acquisitions enter through `read_mrm()`, which reads a Waters DESI-MRM `.raw`
folder and joins the ion library to the features:

```{r read-mrm, eval = FALSE}
obj <- read_mrm(
  name         = "my_acquisition",        # .raw folder name, without the suffix
  folder       = "path/to/raw",           # directory containing my_acquisition.raw
  lib_ion_path = "path/to/ion_library.csv"
)
```

The `.raw` folder must contain the exported per-transition text files (an
`imaging/` subdirectory) or a previously cached `MSImagingExperiment.rds`. The
ion library is a CSV with one row per transition; the columns used for matching
are the precursor and product m/z, plus `transition_id` for the display name and
`Type` to mark internal standards. Any additional columns -- pathway class,
polarity, collision energy -- are carried through to `fData()` and become
available for grouping, as shown in *Joining ion-library metadata* below.

The result is a features x pixels object exactly like the one inspected above,
ready for `select_tissue_pixels()` and the rest of the workflow.

# Loading the acquisitions

*Families: `acquisition`, `combining acquisitions`.*

| Function | Takes | Produces |
|---|---|---|
| `read_mrm()` | `.raw` folder + ion-library CSV | `MSImagingExperiment`, `fData` joined to the library |
| `trim_MSI()` | object with a tissue mask | object with pure-background border rows/columns dropped |
| `remove_blank_mzs()` | object | object with features that have no data removed |
| `setCommonAxis()` | **list** of objects + reference `fData` | list sharing one m/z axis |
| `align_features()` | two objects | both reduced to common features, m/z-keyed |
| `combine_MSIs()` | two or more aligned objects | one object, each acquisition its own `run` |
| `bind_panels()` / `bind_polarities()` | two objects on a shared `(x, y)` grid | features of both on the pixel-grid intersection |

The bundled example is six sections: three of group **A** (circular tissue) and
three of group **B** (square tissue). They ship as RDS, so `read_mrm()` is not
needed here. `combine_MSIs()` binds them into a single object, keeping each
section as its own `run`:

```{r load}
library(quantMSImageR)
library(ggplot2)

sections <- sprintf("section%02d", 1:6)
objs <- lapply(sections, function(s)
  readRDS(system.file("extdata", "example.raw", paste0(s, ".RDS"),
                      package = "quantMSImageR")))
combined <- do.call(combine_MSIs, objs)

groups <- c("A", "A", "A", "B", "B", "B")   # sections 01-03 = A, 04-06 = B
combined
```

Each section contributed its pixels as a separate `run`, on the same feature
axis:

```{r load-check}
dim(combined)                       # features x (pixels from all six sections)
table(pData(combined)$run)
```

Everything below operates on this one `combined` object.

# Selecting tissue pixels

*Family: `acquisition`.*

| Function | Takes | Produces |
|---|---|---|
| `select_tissue_pixels()` | acquisition name, data path, ion library | `tissue_pixels.csv` (`x`, `y`, `tissue_pixels`, `background_pixels`) plus a preview of the saved mask |

Before SNR filtering, each acquisition needs a tissue mask. `select_tissue_pixels()`
displays every feature's ion image so you can pick the one that best separates
tissue from background, then opens an interactive ROI selector:

```{r select, eval = FALSE}
select_tissue_pixels(
  name         = "my_acquisition",          # .raw folder name (no suffix)
  data_path    = "path/to/raw",
  lib_ion_path = "path/to/ion_library.csv"
)
```

It writes the resulting mask as `tissue_pixels.csv` **inside the acquisition's
`.raw` folder**, where `int2snr()` and `generate_txt_images()` read it. The mask
is simply a per-pixel label -- tissue versus background -- which the bundled
sections already carry in their `sample_name` column:

```{r mask-table}
mask_tbl <- as.data.frame(pData(combined))[, c("run", "x", "y", "sample_name")]

DT::datatable(head(mask_tbl, 500), rownames = FALSE,
              options = list(pageLength = 5))
```

**Table 1.** Per-pixel tissue/background labelling used by `int2snr()` (first 500
pixels).

Plotting that column over the pixel coordinates shows the mask itself, which is
the quickest way to confirm a selection before running the pipeline -- the
circular group A and square group B tissue should stand out cleanly from the
surrounding background:

```{r mask-plot, fig.wide = TRUE, fig.cap = "Tissue masks for all six sections. Green pixels are labelled tissue_pixels and contribute to every downstream summary; grey pixels are background_pixels and are used to estimate the noise level in int2snr()."}
ggplot(mask_tbl, aes(x, -y, fill = sample_name)) +
  geom_tile() +
  facet_wrap(~ run, ncol = 3) +
  coord_equal() +
  scale_fill_manual(values = c(tissue_pixels     = "#1b7837",
                               background_pixels = "grey85")) +
  labs(x = NULL, y = NULL, fill = NULL) +
  theme_minimal() +
  theme(axis.text = element_blank(), panel.grid = element_blank(),
        strip.text = element_text(size = 7),
        legend.text = element_text(size = 7))
```

# Ion images

*Family: `visualisation`.*

| Function | Takes | Produces |
|---|---|---|
| `imageR()` | object, `feat_ind`, `val_slot`, `sample_lab`, `scale` | a **`ggplot`** ion image |
| `quantile_hm()` | object, `quant_val`, sample order/labels, `feature_split` | a `ComplexHeatmap::Heatmap` (rows = samples, columns = features) |

`imageR()` returns a `ggplot` ion image for a chosen feature. Because it is a
plain `ggplot`, the layout can be adjusted -- here into a 3 x 2 grid:

```{r image, fig.wide = TRUE, fig.cap = "Ion image of the first feature across all six sections, before SNR filtering. Sections 01-03 are group A (circular tissue), 04-06 are group B (square)."}
imageR(combined, feat_ind = 1, sample_lab = "run", scale = "suppress") +
  facet_wrap(~ sample, ncol = 3) +
  theme(strip.text = element_text(size = 7),
        legend.text = element_text(size = 7))
```

The same per-feature images can be written to disk as `.txt` intensity matrices
(`generate_txt_images(output_txt = TRUE)`, or `output_txt: true` in the YAML
config). Those files open directly in ImageJ, which is the usual route into
multimodal workflows.

# Joining ion-library metadata

`build_feature_meta()` joins ion-library annotation to features by rounded
`(precursor, product)` m/z, so extra columns -- such as the pathway class
`Met-1` -- become available for grouping. We do this before the heatmap because
the heatmap uses `Met-1` to split its rows.

```{r meta}
lib <- read.csv(system.file("extdata", "example_ion_library.csv",
                            package = "quantMSImageR"), check.names = FALSE)
fm <- build_feature_meta(combined, lib)

DT::datatable(fm[, c("name", "precursor_mz", "product_mz", "Type", "Met-1")],
              rownames = FALSE, options = list(pageLength = 10))
```

**Table 2.** Ion-library metadata joined to each feature by m/z.

# Quantile heatmaps across samples

`quantile_hm()` summarises each feature to a per-sample quantile and returns a
`ComplexHeatmap::Heatmap`. **Rows are samples, grouped by study group; columns
are features, split by the `Met-1` class joined above** -- studies usually have
more samples than features, so the long dimension runs vertically. Cells are
drawn square (stretching to at most 1.5:1 when one dimension is much longer),
so the heatmap grows with the data rather than distorting to fill the device.

The `quant_val` argument chooses which quantile summarises each feature -- a
median (`0.5`) reflects the bulk of the tissue, while a high quantile (`0.95`)
emphasises hot-spots:

```{r heatmap-median, fig.wide = TRUE, fig.cap = "Median (50th percentile) feature intensity per section. Rows are sections grouped by study group; columns are features split by Met-1 class."}
feature_split <- fm[["Met-1"]]

quantile_hm(combined, quant_val = 0.5,
            heatmap_order = sections, heatmap_labs = groups,
            feature_split = feature_split, feature_split_name = "Met-1")
```

```{r heatmap-95, fig.wide = TRUE, fig.cap = "The same summary at the 95th percentile, which emphasises hot-spots rather than the bulk of the tissue."}
quantile_hm(combined, quant_val = 0.95,
            heatmap_order = sections, heatmap_labs = groups,
            feature_split = feature_split, feature_split_name = "Met-1")
```

# Signal-to-noise filtering

*Family: `filtering`.*

| Function | Takes | Produces |
|---|---|---|
| `int2response()` | object + internal-standard name | intensities normalised to the IS (`response`) |
| `int2snr()` | object, `background`/`tissue` labels, `snr_thresh`, optional `snr_overrides` | adds an **`snr`** slot; below threshold and non-tissue become `NA` |
| `applySNR()` | object carrying an `snr` slot | sub-threshold pixels set to `NA` in `val_slot` |
| `back2NA()` | object + background label | background pixels set to `NA` |
| `zero2na()` | object, `val_slot` | zeros replaced with `NA` |

`int2snr()` computes a per-pixel signal-to-noise ratio for each feature against
the background pixels, and `applySNR()` then sets every sub-threshold pixel to
`NA` in the intensity slot.

```{r snr-apply}
combined <- int2snr(combined, val_slot = "intensity", snr_thresh = 3)
combined <- applySNR(combined, val_slot = "intensity")
```

`int2snr()` added an `snr` layer rather than replacing anything, and
`applySNR()` then set the sub-threshold pixels of `intensity` to `NA`:

```{r snr-check}
names(spectraData(combined))                            # snr layer added
mean(is.na(spectra(combined, "intensity"))) |> round(3) # fraction now masked
```

Compare the filtered image below with the unfiltered ion images in
*Ion images* above: the low-signal background has been removed, leaving the
tissue.

```{r snr-after, fig.wide = TRUE, fig.cap = "The same feature after applySNR(). Sub-threshold pixels are now NA, so the low-signal background has been removed and only tissue signal remains."}
imageR(combined, feat_ind = 1, sample_lab = "run", val_slot = "intensity",
       scale = "suppress") +
  facet_wrap(~ sample, ncol = 3) +
  theme(strip.text = element_text(size = 7),
        legend.text = element_text(size = 7))
```

# Running a full study from a YAML config

*Family: `workflow`.*

| Function | Takes | Produces |
|---|---|---|
| `run_study()` | a YAML configuration file | runs the whole study: `.txt` images, optional calibration, one HTML report per SNR threshold |
| `generate_txt_images()` | acquisition names, paths, ion library, SNR settings | list of combined/SNR-filtered objects; optional per-feature `.txt` images |
| `run_example()` | nothing (all paths resolved internally) | the same list plus `calibrated` and `report`, having rendered the HTML report |

Everything above runs inside `generate_txt_images()`, which `run_study()` drives
from a YAML file. In practice a whole study -- multiple acquisitions, SNR
sweeps, ion ratios and optional calibration -- is configured in one place. A
minimal config looks like:

```yaml
study: "my_study"
paths:
  data_path:    "path/to/raw"
  out_path:     "path/to/results"
  image_dir:    "path/to/images"
  lib_ion_path: "path/to/ion_library.csv"
samples:
  - pos: "acq_ctrl"
    run_id: "S1"
    label:  "Ctrl"
  - pos: "acq_treated"
    run_id: "S2"
    label:  "Treatment"
parameters:
  snr_thresh: [3]
output:
  render_report: true
  output_txt:    true      # write per-feature .txt images for ImageJ
```

The shipped template documents every option (SNR overrides, ratios,
calibration, ...) -- open it to use as a starting point:

```{r yaml-template, eval = FALSE}
file.show(system.file("config_template.yaml", package = "quantMSImageR"))
```

Then point `run_study()` at your filled-in config:

```{r run-study, eval = FALSE}
run_study("path/to/study_config.yaml")
```

The same runner is available from the command line, which is the usual way to
run a large study unattended:

```sh
Rscript -e 'quantMSImageR::run_study("study_config.yaml")'
```

# References

Smith MJ, *et al.* Quantitative mass spectrometry imaging of lipid mediators
using DESI-MRM. *Analytical Chemistry* (2024).
<https://pubs.acs.org/doi/10.1021/acs.analchem.4c02350>

# Session information

```{r session}
sessionInfo()
```
