---
title: "Introduction to quantMSImageR"
author: "Matthew J. Smith"
date: "`r Sys.Date()`"
package: quantMSImageR
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Introduction to quantMSImageR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE,
                      message = FALSE)
```

# Introduction

`quantMSImageR` extends
[Cardinal](https://bioconductor.org/packages/Cardinal/) with tools for
processing and **quantifying** targeted DESI-MRM mass spectrometry imaging (MSI)
data: signal-to-noise (SNR) filtering, tissue/background separation,
per-feature ion images, and quantile heatmaps.

This vignette walks through the core imaging workflow on the small synthetic
dataset bundled with the package. Absolute quantification against calibration
standards is covered separately in the *Absolute quantification with
calibration standards* vignette
(`vignette("quantification", package = "quantMSImageR")`).

Every acquisition is a `Cardinal::MSImagingExperiment`; most `quantMSImageR`
functions operate on the `quant_MSImagingExperiment` subclass, so we coerce
first.

# Installation

```{r install, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("quantMSImageR")
```

# Quick start

The fastest way to see the whole pipeline is `run_example()`, which runs the
imaging workflow on two synthetic groups (**A**, circular tissue, n = 3;
**B**, square tissue, n = 3), renders the HTML report, and also demonstrates
absolute quantification on the bundled calibration standards:

```{r quickstart, eval = FALSE}
library(quantMSImageR)
res <- run_example()          # both groups; imaging report + calibration demo
res$calibrated                # the calibrated object
```

The rest of this vignette reproduces the key steps by hand.

# Loading an acquisition

```{r load}
library(quantMSImageR)

p <- system.file("extdata", "example.raw", "section01.RDS",
                 package = "quantMSImageR")
obj <- as(readRDS(p), "quant_MSImagingExperiment")
obj
```

# Selecting tissue pixels

Before SNR filtering, each acquisition needs a tissue mask. `select_tissue_pixels()`
displays every feature's ion image so you can pick the one that best separates
tissue from background, then opens an interactive ROI selector:

```{r select, eval = FALSE}
select_tissue_pixels(
  name         = "my_acquisition",     # .raw folder name (no suffix)
  data_path    = "D:/studies/raw",
  lib_ion_path = "D:/studies/ion_library.csv"
)
```

It writes the resulting logical mask as `tissue_pixels.csv` **inside the
acquisition's `.raw` folder**, where `int2snr()` (below) and
`generate_txt_images()` read it to label `tissue_pixels` vs `noise_pixels`. The
bundled sections already carry this mask embedded in their `sample_name` column,
so the steps below run without it.

# Ion images

`imageR()` returns a `ggplot` ion image for a chosen feature:

```{r image, fig.small = TRUE}
imageR(obj, feat_ind = 1, sample_lab = "run", scale = "suppress")
```

# Quantile heatmaps across samples

`quantile_hm()` summarises each feature to a per-sample quantile and returns a
`ComplexHeatmap::Heatmap`, with columns grouped by study group. We combine the
six bundled sections (three per group) into one object with `combine_MSIs()`:

```{r combine}
sections <- sprintf("section%02d", 1:6)
objs <- lapply(sections, function(s)
  readRDS(system.file("extdata", "example.raw", paste0(s, ".RDS"),
                      package = "quantMSImageR")))
combined <- do.call(combine_MSIs, objs)

groups <- c("A", "A", "A", "B", "B", "B")   # sections 01-03 = A, 04-06 = B
```

The `quant_val` argument chooses which quantile summarises each feature -- a
median (`0.5`) reflects the bulk of the tissue, while a high quantile (`0.95`)
emphasises hot-spots:

```{r heatmap-median, fig.small = TRUE}
quantile_hm(combined, quant_val = 0.5,
            heatmap_order = sections, heatmap_labs = groups)
```

```{r heatmap-95, fig.small = TRUE}
quantile_hm(combined, quant_val = 0.95,
            heatmap_order = sections, heatmap_labs = groups)
```

# Signal-to-noise filtering

`int2snr()` computes a per-pixel signal-to-noise ratio for each feature against
the noise pixels, and `applySNR()` then sets every sub-threshold pixel to `NA`
in the intensity slot. Compare feature 1 before and after: the low-signal
background is removed, leaving the tissue.

```{r snr-before, fig.small = TRUE}
# before -- raw intensity (background still present)
imageR(obj, feat_ind = 1, sample_lab = "run", val_slot = "intensity",
       scale = "suppress")
```

```{r snr-apply}
obj <- int2snr(obj, val_slot = "intensity", sample_type = "sample_name",
               noise = "noise_pixels", tissue = "tissue_pixels", snr_thresh = 3)
obj <- applySNR(obj, val_slot = "intensity")
```

```{r snr-after, fig.small = TRUE}
# after -- pixels with SNR < 3 are masked out
imageR(obj, feat_ind = 1, sample_lab = "run", val_slot = "intensity",
       scale = "suppress")
```

# Joining ion-library metadata

`build_feature_meta()` joins ion-library annotation to features by rounded
`(precursor, product)` m/z, so extra columns (e.g. a pathway class) can drive
row grouping in reports.

```{r meta}
lib <- read.csv(system.file("extdata", "example_ion_library.csv",
                            package = "quantMSImageR"), check.names = FALSE)
fm <- build_feature_meta(combined, lib)
head(fm[, c("name", "Met-1")])
```

# Running a full study from a YAML config

In practice a whole study -- multiple acquisitions, SNR sweeps, fold-change
layers, ion ratios and optional calibration -- is driven from a single YAML
configuration. A minimal config looks like:

```yaml
study: "my_study"
paths:
  data_path:    "D:/study/raw"
  out_path:     "D:/study/results"
  image_dir:    "D:/study/images"
  lib_ion_path: "D:/study/ion_library.csv"
samples:
  - pos: "acq_ctrl"
    run_id: "S1"
    label:  "Ctrl"
  - pos: "acq_treated"
    run_id: "S2"
    label:  "Treatment"
parameters:
  snr_thresh: [3]
output:
  render_report: true
```

The shipped template documents every option (SNR overrides, ratios,
calibration, …) -- open it to use as a starting point:

```{r yaml-template, eval = FALSE}
file.show(system.file("config_template.yaml", package = "quantMSImageR"))
```

Point `CONFIG_FILE` at your filled-in config and source the runner:

```{r run-study, eval = FALSE}
CONFIG_FILE <- "path/to/study_config.yaml"
source(system.file("run_study.R", package = "quantMSImageR"))
```

# Session info

```{r session}
sessionInfo()
```
