---
title: "Quantification of MSI data with quantMSImageR"
author: "Matthew J. Smith"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quantification of MSI data with quantMSImageR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", warning = FALSE,
                      message = FALSE)

# Whether the synthetic calibration data has been generated (gates the
# calibration chunk). Defined here so it is always available.
cal_dir  <- system.file("extdata", "cal_example.raw", package = "quantMSImageR")
have_cal <- nzchar(cal_dir) && file.exists(file.path(cal_dir, "cal_MSI.RDS"))
```

# Introduction

`quantMSImageR` extends [Cardinal](https://bioconductor.org/packages/Cardinal/)
with tools for processing and **quantifying** targeted DESI-MRM mass
spectrometry imaging (MSI) data: signal-to-noise (SNR) filtering,
tissue/background separation, per-feature ion images, quantile heatmaps, and
standards-based absolute quantification.

This vignette walks through the core workflow on the small synthetic dataset
bundled with the package. Every acquisition is a
`Cardinal::MSImagingExperiment`; most `quantMSImageR` functions operate on the
`quant_MSImagingExperiment` subclass, so we coerce first.

```{r load}
library(quantMSImageR)

p <- system.file("extdata", "example.raw", "section01.RDS",
                 package = "quantMSImageR")
obj <- as(readRDS(p), "quant_MSImagingExperiment")
obj
```

# Signal-to-noise filtering

`int2snr()` computes a per-pixel signal-to-noise ratio for each feature against
the noise pixels, and `applySNR()` masks intensities that fall below the
threshold. The bundled sections carry an `intensity` slot and a `sample_name`
column labelling `tissue_pixels` / `noise_pixels`.

```{r snr}
obj <- int2snr(obj, val_slot = "intensity", sample_type = "sample_name",
               noise = "noise_pixels", tissue = "tissue_pixels", snr_thresh = 3)
obj <- applySNR(obj, val_slot = "intensity")
```

# Ion images and quantile heatmaps

`imageR()` returns a `ggplot` ion image for a chosen feature; `quantile_hm()`
summarises each feature to a per-sample quantile and returns a
`ComplexHeatmap::Heatmap`.

```{r image, fig.small = TRUE}
imageR(obj, feat_ind = 1, sample_lab = "run", scale = "suppress")
```

```{r heatmap, fig.small = TRUE}
hm <- quantile_hm(obj, quant_val = 0.5,
                  heatmap_order = "section01", heatmap_labs = "A")
hm
```

# Joining ion-library metadata

`build_feature_meta()` joins ion-library annotation to features by rounded
`(precursor, product)` m/z, so extra columns (e.g. a pathway class) can drive
row grouping in reports.

```{r meta}
lib <- read.csv(system.file("extdata", "example_ion_library.csv",
                            package = "quantMSImageR"), check.names = FALSE)
fm <- build_feature_meta(obj, lib)
head(fm[, c("name", "Met-1")])
```

# Absolute quantification with calibration standards

When a calibration standards acquisition is available, `quantMSImageR` fits a
response-versus-amount curve per lipid and converts tissue-pixel intensities to
`pg/pixel` and `pg/mm2`. The chain is
`summarise_cal_levels()` -> `create_cal_curve()` -> `int2conc()`.

The synthetic calibration dataset is produced by
`inst/generate_cal_data.R`; the chunk below runs only when it is present.

```{r calibration, eval = have_cal}
cal  <- as(readRDS(file.path(cal_dir, "cal_MSI.RDS")),
           "quant_MSImagingExperiment")
cal_metadata <- read.csv(file.path(cal_dir, "calibration_metadata.csv"))

cal <- summarise_cal_levels(cal, cal_metadata, val_slot = "intensity",
                            cal_label = "Cal", id = "identifier")
cal <- create_cal_curve(cal, cal_type = "cal")

# R-squared per lipid curve
cal@calibrationInfo@r2_df

# convert tissue pixels to concentration
cal <- int2conc(cal, val_slot = "intensity", pixels = "Tissue")
range(spectra(cal, "conc - pg/mm2"), na.rm = TRUE)
```

In a full study this calibration is driven from the YAML configuration (a
`calibration:` block with the standards acquisition, ROI file and metadata),
run via `source(system.file("run_study.R", package = "quantMSImageR"))`.

# Session info

```{r session}
sessionInfo()
```
