---
title: "MSI study report"
author: "Matt Smith"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    html_document :
      code_folding: hide
      number_sections: false
      toc: true
      toc_float: 
        collapsed: false
        smooth_scroll: true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# set default options
knitr::opts_chunk$set(echo = FALSE,
               message = FALSE,
               warning = FALSE,
               fig.align="center",
               dpi = 120,
               fig.retina = 3)

# Define a function to safely plot or show a blank plot
safe_plot <- function(p) {
  tryCatch({
    plot(p)
  }, error = function(e) {
    message("Plot failed: ", e$message)
    plot.new()
    title("Blank plot (error occurred)")
  })
}

make_sim_mat <- function(out, metric = "Dice") {
  mets <- sort(unique(c(out$metA, out$metB)))
  
  # wide upper-triangle -> matrix
  w <- out %>%
    select(metA, metB, value = all_of(metric)) %>%
    filter(metA != metB) %>%
    mutate(metA = factor(metA, levels = mets),
           metB = factor(metB, levels = mets)) %>%
    tidyr::complete(metA, metB) %>%   # ensure full square
    pivot_wider(names_from = metB, values_from = value)
  
  mat <- as.matrix(w[,-1])
  rownames(mat) <- as.character(w$metA)
  
  # symmetrize (upper/lower may be NA depending on how 'out' was built)
  mat_t <- t(mat)
  mat_sym <- ifelse(is.na(mat), mat_t, mat)
  mat_sym <- ifelse(is.na(mat_sym), mat, mat_sym)
  
  diag(mat_sym) <- 1
  mat_sym
}

cluster_from_sim <- function(sim_mat, method = "average") {
  # handle any remaining NA (choose one strategy)
  # option A: treat missing as 0 similarity
  sim_mat[is.na(sim_mat)] <- 0
  
  d <- as.dist(1 - sim_mat)
  hclust(d, method = method)
}

plot_tri_heatmap <- function(out, metric = "Dice", ord) {
  
  tmp <- out %>%
    select(metA, metB, value = all_of(metric)) %>%
    filter(metA != metB) %>%
    mutate(
      metA = factor(metA, levels = ord),
      metB = factor(metB, levels = ord),
      i = as.integer(metA),
      j = as.integer(metB)
    ) %>%
    filter(j > i)
  
  ggplot(tmp, aes(x = metA, y = metB, fill = value)) +
    geom_tile(color = "grey85", linewidth = 0.1) +
    scale_x_discrete(limits = ord, drop = FALSE) +
    scale_y_discrete(limits = ord, drop = FALSE) +
    scale_fill_gradientn(
      colours = c("#053061","#92C5DE","white","#F4A582","#67001F"),
      limits = c(min(tmp$value, na.rm = TRUE)-0.1, max(tmp$value, na.rm = TRUE)+0.1),
      na.value = "white"
    ) +
    labs(fill = "Colocalization") +
    coord_fixed() +
    theme_minimal(base_size = 10) +
    theme(
      axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
      axis.title = element_blank(),
      panel.grid = element_blank(),
      legend.position = "right"
    )
}

```


# 1. Study

DESI-MRM mass spectrometry imaging of tissue sections, processed with quantMSImageR.

## 1.1 Sample mapping

Acquisition filenames, run IDs used throughout this report, and group labels.
Where a group contains biological replicates, each section is assigned a unique
`_1`, `_2`, … suffix (shown individually in heatmaps and ion images; pooled by
group label in comparison plots).

```{r sample-map, echo = FALSE}
if (exists("sample_map")) {
  # Hide empty polarity columns for cleaner display
  sm_show <- sample_map
  if (all(sm_show$pos_files == "")) sm_show$pos_files <- NULL
  if (all(sm_show$neg_files == "")) sm_show$neg_files <- NULL
  DT::datatable(sm_show, rownames = FALSE,
                options = list(pageLength = 25, scrollX = TRUE))
} else {
  cat("*(sample_map not available — run via run_study.R to populate)*")
}
```

## 1.2 Features / transitions

MRM transitions and analyte metadata from the ion library.

```{r import, echo = FALSE}

DT::datatable(data.frame(fData(combined)))

```

# 2. Heatmap {.tabset .tabset-fade}

Z-scored nth-percentile intensity per feature across samples (clipped to [−1, 1]).
Rows are features; columns are individual sections ordered by group.

```{r hm, results="asis", echo = FALSE, include=T, warning=FALSE, fig.width=7.5, fig.height=6}

for(quant_value in c(0.25, 0.5, 0.75, 0.95, 0.98)){
  
  cat( paste('## ', as.character(quant_value*100), "th quantile heatmap", '\n' ) )
  
  p = quantile_hm(MSIobject = combined,
                  quant_val = quant_value,
                  heatmap_order = heatmap_order,
                  heatmap_labs = heatmap_labs)
  
  print(p)  # Explicitly print the heatmap object
  
  cat('\n \n')
   
}

```

```{r cmp-heading, results="asis", echo=FALSE}
if (length(heatmap_order) > 1)
  cat("# 3. Comparison plots {.tabset .tabset-fade}\n\n## Box + violin (side-by-side)\n\n")
```

```{r box_violin_dropdown, results="asis", echo=FALSE, include=TRUE, warning=FALSE, message=FALSE, fig.width=11, fig.height=5.5, eval=length(heatmap_order) > 1}

val_slot <- "intensity"

stopifnot(length(heatmap_order) == length(heatmap_labs))

# Map run -> group label (can repeat)
run_to_lab <- setNames(heatmap_labs, heatmap_order)

# Unique display labels for plotting (avoid duplicated factor levels)
`%||%` <- function(a, b) if (is.null(a)) b else a

make_unique_labels <- function(labs) {
  out <- character(length(labs))
  seen <- list()
  for (i in seq_along(labs)) {
    x <- labs[i]
    k <- (seen[[x]] %||% 0) + 1
    seen[[x]] <- k
    if (sum(labs == x) > 1) out[i] <- sprintf("%s (%d)", x, k) else out[i] <- x
  }
  out
}

plot_labs <- make_unique_labels(heatmap_labs)
run_to_plotlab <- setNames(plot_labs, heatmap_order)
plot_levels <- plot_labs  # already in heatmap_order order, unique

# Baseline label from calling script if present; else default to first label
if (!exists("baseline_label") || is.null(baseline_label) || !nzchar(baseline_label)) {
  baseline_label <- heatmap_labs[1]
}
if (!(baseline_label %in% heatmap_labs)) {
  stop(sprintf("baseline_label '%s' not found in heatmap_labs", baseline_label))
}

# feature names for titles
met_names <- tryCatch(featureNames(combined), error = function(e) NULL)
if (is.null(met_names)) met_names <- as.character(mz(combined))

# safe ids for dropdown blocks
make_safe_id <- function(x) gsub("[^A-Za-z0-9_\\-]", "_", x)
safe_ids <- vapply(met_names, make_safe_id, character(1))

# ---- Detect features with any signal in selected runs ----
is_detected <- vapply(seq_along(met_names), function(mz_ind) {
  any(vapply(heatmap_order, function(fn) {
    p <- combined$run == fn
    v <- spectraData(combined)[[val_slot]][mz_ind, p]
    any(is.finite(v)) && any(v[is.finite(v)] != 0)
  }, logical(1)))
}, logical(1))

not_detected  <- met_names[!is_detected]
detected_mets <- met_names[is_detected]
detected_inds <- which(is_detected)
detected_ids  <- safe_ids[is_detected]

# Alphabetical dropdown
ord <- order(tolower(detected_mets))
detected_mets <- detected_mets[ord]
detected_inds <- detected_inds[ord]
detected_ids  <- detected_ids[ord]

# ---- Dropdown + JS ----
cat('
<label for="cmp_select" style="font-weight:600;">Select metabolite:</label>
<select id="cmp_select" style="margin: 0 0 12px 8px; padding: 4px 6px;"
        onchange="showCmp(this.value)">
')

if (length(detected_mets) == 0) {
  cat('<option value="">No detected metabolites</option>')
} else {
  for (i in seq_along(detected_mets)) {
    sel <- if (i == 1) " selected" else ""
    cat(sprintf('<option value="%s"%s>%s</option>\n', detected_ids[i], sel, detected_mets[i]))
  }
}

cat('</select>

<script>
function showCmp(id){
  var blocks = document.querySelectorAll(".cmp_block");
  for (var i=0; i<blocks.length; i++) { blocks[i].style.display = "none"; }
  var el = document.getElementById("cmp_" + id);
  if (el) el.style.display = "block";
}
</script>
')

# ============================
# Collect FC rows during plotting
# FC computed by GROUP label ("Ctrl", "HDM"), pooling runs per group
# ============================
fc_rows <- list()

if (length(detected_mets) > 0) {
  
  for (i in seq_along(detected_mets)) {
    
    mz_ind <- detected_inds[i]
    met    <- detected_mets[i]
    mid    <- detected_ids[i]
    
    # Build long df with plot label (unique per run) and group label (pooled for FC)
    df <- bind_rows(lapply(heatmap_order, function(fn) {
      p <- combined$run == fn
      v <- spectraData(combined)[[val_slot]][mz_ind, p]
      v <- v[is.finite(v)]
      tibble(
        run = fn,
        plot_label  = run_to_plotlab[[fn]],
        group_label = run_to_lab[[fn]],
        intensity = v
      )
    }))
    
    if (nrow(df) == 0) next
    
    # factor for plotting (unique, ordered)
    df$plot_label <- factor(df$plot_label, levels = plot_levels)
    # group factor pools replicates (e.g. HDM_M_1 + HDM_M_2 → HDM_M) for boxplots
    df$group_label <- factor(df$group_label, levels = unique(heatmap_labs))
    
    # ---- FC computation: quantiles per GROUP label ----
    q_tbl <- df %>%
      group_by(group_label) %>%
      summarise(
        q25 = unname(quantile(intensity, 0.25, na.rm = TRUE)),
        q50 = unname(quantile(intensity, 0.50, na.rm = TRUE)),
        q75 = unname(quantile(intensity, 0.75, na.rm = TRUE)),
        q90 = unname(quantile(intensity, 0.90, na.rm = TRUE)),
        .groups = "drop"
      )
    
    base_row <- q_tbl %>% filter(group_label == baseline_label)
    
    if (nrow(base_row) == 1) {
      
      base_q25 <- base_row$q25
      base_q50 <- base_row$q50
      base_q75 <- base_row$q75
      base_q90 <- base_row$q90
      
      if (all(is.finite(c(base_q25, base_q50, base_q75, base_q90))) &&
          all(c(base_q25, base_q50, base_q75, base_q90) != 0)) {
        
        fc_one <- q_tbl %>%
          filter(group_label != baseline_label) %>%
          transmute(
            metabolite = met,
            
            # FC (requested order)
            FC_q50 = q50 / base_q50,
            FC_q90 = q90 / base_q90,
            FC_q75 = q75 / base_q75,
            FC_q25 = q25 / base_q25,
            
            # log2FC (same order)
            log2FC_q50 = log2(q50 / base_q50),
            log2FC_q90 = log2(q90 / base_q90),
            log2FC_q75 = log2(q75 / base_q75),
            log2FC_q25 = log2(q25 / base_q25),
            
            baseline = baseline_label,
            compare  = as.character(group_label),
            
            # the rest (QC/debugging)
            baseline_q25 = base_q25,
            baseline_q50 = base_q50,
            baseline_q75 = base_q75,
            baseline_q90 = base_q90,
            
            compare_q25 = q25,
            compare_q50 = q50,
            compare_q75 = q75,
            compare_q90 = q90
          )
        
        fc_rows[[length(fc_rows) + 1]] <- fc_one
      }
    }
    
    # ---- Visualization clipping (global p90 across pixels shown) ----
    p90_vis  <- quantile(df$intensity, 0.90, na.rm = TRUE)
    pmin_vis <- min(df$intensity, na.rm = TRUE)
    
    p_box <- ggplot(df, aes(x = group_label, y = intensity, fill = group_label)) +
      theme_classic() +
      geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.7) +
      coord_cartesian(ylim = c(pmin_vis, p90_vis)) +
      labs(x = NULL, y = "Intensity (≤ p90)") +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
            legend.position = "none")

    p_violin <- ggplot(df, aes(x = group_label, y = intensity, fill = group_label)) +
      theme_classic() +
      geom_violin(trim = TRUE, alpha = 0.6) +
      geom_boxplot(width = 0.15, outlier.shape = NA, alpha = 0.8) +
      coord_cartesian(ylim = c(pmin_vis, p90_vis)) +
      labs(x = NULL, y = "Intensity (≤ p90)") +
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
            legend.position = "none")
    
    p_both <- (p_box + ggtitle("Boxplot")) | (p_violin + ggtitle("Violin + box")) +
      patchwork::plot_annotation(title = met)
    
    display_style <- if (i == 1) "block" else "none"
    cat(sprintf('<div class="cmp_block" id="cmp_%s" style="display:%s;">\n', mid, display_style))
    print(p_both)
    cat("</div>\n")
  }
}

# ---- Not-detected list (QC) ----
if (length(not_detected) > 0) {
  cat("<details style='margin-top:12px;'><summary><b>Not detected metabolites</b></summary><ul>")
  for (x in not_detected) cat(sprintf("<li>%s</li>", x))
  cat("</ul></details>")
}

# ---- FC table (built during plotting) ----
fc_tbl <- dplyr::bind_rows(fc_rows)

cat("<hr><h3>Fold change table (quantiles)</h3>")

if (nrow(fc_tbl) == 0) {
  cat("<p><b>No fold-changes computed.</b> Likely baseline quantiles were 0/NA for many metabolites, or one group had no pixels.</p>")
} else {
  
  # requested column order first, then keep the rest
  wanted <- c(
    "metabolite",
    "FC_q50", "FC_q90", "FC_q75", "FC_q25",
    "log2FC_q50", "log2FC_q90", "log2FC_q75", "log2FC_q25",
    "baseline", "compare"
  )
  keep_first <- intersect(wanted, colnames(fc_tbl))
  fc_tbl <- fc_tbl %>% dplyr::select(dplyr::all_of(keep_first), dplyr::everything())
  
  cat(sprintf(
    "<p style='margin-top:-6px;'>Baseline: <b>%s</b>. FC computed as compare/baseline using Q25, Q50 (median), Q75, and Q90 of pooled pixel intensities.</p>",
    baseline_label
  ))
  
  DT::datatable(
    fc_tbl,
    options = list(pageLength = 25, scrollX = TRUE),
    rownames = FALSE
  )
}

```

# 4. Correlation {.tabset .tabset-fade}

Spatial co-localisation metrics computed pairwise across all features.

- **Pearson's r** — linear relationship between pixel intensities of two ions.
- **Manders Overlap Coefficient (MOC)** — degree of pixel-intensity overlap; M1 is the fraction of ion 1 intensity that overlaps with ion 2, M2 the reverse (asymmetric).
- **Dice score** — overlap of the binary segmentation masks derived from M1 and M2.

Metrics are reported for all samples pooled (**All**) and for each group separately, allowing differences in co-localisation structure between groups to be identified.

```{r cor-compute, echo=FALSE, include=FALSE, warning=FALSE, message=FALSE}

# Metabolite names (use featureNames if available; fallback to mz labels)
met_names <- tryCatch(featureNames(combined), error = function(e) NULL)
if (is.null(met_names)) met_names <- as.character(mz(combined))
mz_vals   <- mz(combined)
name_to_idx <- setNames(seq_along(met_names), met_names)

# --- Helper: compute per-pair correlation long table for a given MSI object ---
compute_coloc_table <- function(obj) {
  if (ncol(obj) == 0) return(NULL)
  raw <- do.call(
    rbind,
    lapply(seq_along(mz_vals), function(j) {
      tryCatch({
        tmp <- data.frame(Cardinal::colocalized(obj, mz = mz_vals[j]),
                          check.names = FALSE)
        data.frame(
          metA = met_names[j],
          metB = rownames(tmp),
          cor  = tmp$cor,
          MOC  = tmp$MOC,
          M1   = tmp$M1,
          M2   = tmp$M2,
          Dice = tmp$Dice,
          stringsAsFactors = FALSE,
          check.names = FALSE
        )
      }, error = function(e) NULL)
    })
  )
  if (is.null(raw) || nrow(raw) == 0) return(NULL)
  raw <- raw |>
    dplyr::mutate(
      idxA = unname(name_to_idx[metA]),
      idxB = unname(name_to_idx[metB])
    ) |>
    dplyr::filter(!is.na(idxA), !is.na(idxB), idxA != idxB)
  pairs <- raw |>
    dplyr::mutate(
      idx1 = pmin(idxA, idxB),
      idx2 = pmax(idxA, idxB),
      met1 = met_names[idx1],
      met2 = met_names[idx2],
      pair_id = paste0(idx1, "__", idx2)
    ) |>
    dplyr::filter(idx1 < idx2) |>
    dplyr::distinct(pair_id, .keep_all = TRUE) |>
    dplyr::select(metA = met1, metB = met2, cor, MOC, M1, M2, Dice,
                  idxA = idx1, idxB = idx2, pair_id) |>
    dplyr::filter(is.finite(cor), is.finite(MOC),
                  is.finite(M1), is.finite(M2), is.finite(Dice))
  list(long = raw, pairs = pairs)
}

# --- Build one correlation dataset per group + one for "All" ---
cor_datasets <- list()
cor_datasets[["All"]] <- compute_coloc_table(combined)

unique_groups <- unique(heatmap_labs)
if (length(unique_groups) > 1) {
  for (grp in unique_groups) {
    grp_runs <- heatmap_order[heatmap_labs == grp]
    grp_obj  <- combined[, pData(combined)$run %in% grp_runs]
    cor_datasets[[grp]] <- compute_coloc_table(grp_obj)
  }
}
# Drop failed groups
cor_datasets <- cor_datasets[!vapply(cor_datasets, is.null, logical(1))]

```

## 4.1 Correlation table {.tabset .tabset-fade}

```{r cor-tbl, results="asis", echo = FALSE, include=T, warning=FALSE, fig.width=7.5, fig.height=6}

for (grp_name in names(cor_datasets)) {
  cat("### ", grp_name, "\n\n", sep = "")
  tbl <- cor_datasets[[grp_name]]$pairs
  print(htmltools::tagList(
    DT::datatable(tbl, options = list(pageLength = 15, scrollX = TRUE))
  ))
  cat("\n\n")
}

```

## 4.2 Correlation heat map {.tabset .tabset-fade}

```{r cor-hm, results="asis", echo = FALSE, include=T, warning=FALSE, fig.width=7.5, fig.height=6}

for (grp_name in names(cor_datasets)) {

  cat("### ", grp_name, " {.tabset .tabset-fade}\n\n", sep = "")

  long_df <- cor_datasets[[grp_name]]$long

  for (cor_type in c("cor", "MOC", "M1", "M2", "Dice")) {

    cat("#### ", cor_type, "\n\n", sep = "")

    sim_mat <- tryCatch(make_sim_mat(long_df, cor_type), error = function(e) NULL)
    if (is.null(sim_mat) || nrow(sim_mat) < 2) {
      cat("*(insufficient data)*\n\n"); next
    }
    hc  <- cluster_from_sim(sim_mat)
    ord <- hc$labels[hc$order]

    p <- plot_tri_heatmap(long_df, cor_type, ord)
    print(p)
    cat("\n\n")
  }
}

```

# 5. Ion images {.tabset .tabset-fade}

Ion images of all tissue sections. Three rendering modes per feature:

1. **Histogram normalised** — contrast enhanced per section; best for visualising spatial distribution within a section.
2. **Raw intensities** — no normalisation or scaling; dominated by hotspots but preserves absolute differences between sections.
3. **Global 95th-percentile scale** — intensities capped at the p95 across all sections; balances spatial detail with cross-section intensity comparability.

```{r ion-img-dropdown, results="asis", echo=FALSE, include=TRUE, warning=FALSE, fig.width=7.5, fig.height=6}
# ---- Helper to make safe HTML ids ----
make_safe_id <- function(x) gsub("[^A-Za-z0-9_\\-]", "_", x)

features <- as.character(fData(combined)$name)
feature_ids <- vapply(features, make_safe_id, character(1))

ord <- order(tolower(features))

# sorted display vectors
features_sorted    <- features[ord]
feature_ids_sorted <- feature_ids[ord]

# ---- Dropdown + JS ----
cat('
<label for="ion_select" style="font-weight:600;">Select feature:</label>
<select id="ion_select" style="margin: 0 0 12px 8px; padding: 4px 6px;"
        onchange="showIonFeature(this.value)">
')

for (i in seq_along(features_sorted)) {
  sel <- if (i == 1) ' selected' else ''
  cat(sprintf('<option value="%s"%s>%s</option>\n',
              feature_ids_sorted[i], sel, features_sorted[i]))
}

cat('</select>

<script>
function showIonFeature(id){
  var blocks = document.querySelectorAll(".ion_feature_block");
  for (var i=0; i<blocks.length; i++) { blocks[i].style.display = "none"; }
  var el = document.getElementById("ion_" + id);
  if (el) el.style.display = "block";
}
</script>
')

# ---- Render each feature block (hidden except first) ----
for (i in seq_along(ord)) {

  idx <- ord[i]  # <-- original row index in combined/fData/spectraData

  feature <- features_sorted[i]
  fid     <- feature_ids_sorted[i]
  mz_val  <- fData(combined)$mz[idx]

  display_style <- if (i == 1) "block" else "none"
  cat(sprintf('<div class="ion_feature_block" id="ion_%s" style="display:%s;">\n',
              fid, display_style))
  # Use a styled <p> instead of <h3> so the feature name does not populate the TOC
  cat(sprintf("<p style='margin-top:10px;font-size:1.17em;font-weight:600;'>%s</p>\n", feature))

  # q95 from non-NA pixels only; as.numeric() forces out of matter-array to plain R vector
  all_ints <- as.numeric(spectraData(combined[idx, ])[["intensity"]])
  q95      <- suppressWarnings(quantile(all_ints, 0.95, na.rm = TRUE))
  has_signal <- any(is.finite(all_ints))   # FALSE when feature absent in every sample

  # Helper: blank grid page with message (keeps device in grid mode after Cardinal's grid rendering)
  blank_grid <- function(msg) {
    grid::grid.newpage()
    grid::grid.text(msg, gp = grid::gpar(fontsize = 11, col = "grey40"))
  }

  # Plot 1: Histogram normalization & scaled
  if (!has_signal) {
    blank_grid("Histogram normalization & scaled\n(No signal in any sample)")
  } else {
    tryCatch({
      p1 <- image(combined, enhance = "histogram", scale = TRUE, mz = mz_val,
                  free = "xy", axes = FALSE, grid = FALSE) |>
        matter::set_title("Histogram normalization & scaled")
      plot(p1)
    }, error = function(e) {
      blank_grid("Histogram normalization & scaled\n(No signal or error)")
    })
  }

  # Plot 2: Scaled to 95th quantile
  # Subset to single feature, convert matter-backed array to plain matrix, cap at q95,
  # replace entire slot (sub-slice assignment does not persist in matter backend).
  if (!has_signal) {
    blank_grid("Scaled to 95th quantile\n(No signal in any sample)")
  } else {
    tryCatch({
      combined_q95 <- combined[idx, ]
      ints_q95 <- as.matrix(spectraData(combined_q95)[["intensity"]])
      if (is.finite(q95)) ints_q95[!is.na(ints_q95) & ints_q95 > q95] <- q95
      spectraData(combined_q95)[["intensity"]] <- ints_q95
      p2 <- image(combined_q95, scale = FALSE, zlim = c(0, q95), mz = mz_val,
                  free = "xy", axes = FALSE, grid = FALSE) |>
        matter::set_title("Scaled to 95th quantile")
      plot(p2)
    }, error = function(e) {
      blank_grid("Scaled to 95th quantile\n(No signal or error)")
    })
  }

  # Plot 3: No scaling or normalization (global colorscale across all samples)
  # Same pattern: subset → plain matrix → replace full slot → scale=FALSE + zlim.
  if (!has_signal) {
    blank_grid("No scaling or normalization\n(No signal in any sample)")
  } else {
    tryCatch({
      combined_p3 <- combined[idx, ]
      ints_p3 <- as.matrix(spectraData(combined_p3)[["intensity"]])
      glim <- max(ints_p3, na.rm = TRUE)
      spectraData(combined_p3)[["intensity"]] <- ints_p3
      p3 <- image(combined_p3, scale = FALSE, zlim = c(0, glim), mz = mz_val,
                  free = "xy", axes = FALSE, grid = FALSE) |>
        matter::set_title("No scaling or normalization")
      plot(p3)
    }, error = function(e) {
      blank_grid("No scaling or normalization\n(No signal or error)")
    })
  }

  cat("</div>\n")
}
```

# 6. Session Information

Packages used for the creation of this file

```{r session, echo = FALSE, warning=FALSE}

sessionInfo()

```


